---
title: "Assignment: Colorado Temperature Analysis"
subtitle: "Geospatial Analysis of Climate Change in Colorado"
author: "Your Name Here"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 6)
```

## Instructions

In this assignment, you will analyze temperature data for Colorado using the *same techniques* we learned with precipitation data. You have been provided with two temperature raster files:

- `co_historic_temp.tiff` - Historic mean annual temperature (1970-2000) in degrees Celsius
- `co_2050_temp.tiff` - Projected mean annual temperature (2041-2060) in degrees Celsius

Complete the following tasks using R and the geospatial packages we learned about in class.

---

## Setup

Make sure you have loaded all necessary packages:

```{r load-packages}
library(tidyverse)
library(sf)
library(terra)
library(mapview)
```

Load the spatial objects we used in class:

```{r load-spatial-data}
# Load Colorado boundary data
colorado <- st_read("data/colorado.shp")

# Load Colorado county boundaries
co_counties <- st_read("data/co_counties.shp")

# Create cities dataset
colorado_cities <- data.frame(
  city_name = c("Denver", "Colorado Springs", "Aurora", "Fort Collins",
                "Lakewood", "Thornton", "Arvada", "Pueblo",
                "Grand Junction", "Boulder"),
  longitude = c(-104.9903, -104.8214, -104.8319, -105.0844,
                -105.0769, -104.9719, -105.0875, -104.6091,
                -108.5506, -105.2705),
  latitude = c(39.7392, 38.8339, 39.7294, 40.5853,
               39.7047, 39.8680, 39.8028, 38.2544,
               39.0639, 40.0150),
  population_thousands = c(715.5, 478.2, 366.6, 164.2,
                           155.9, 136.2, 118.4, 111.1,
                           63.3, 105.1)
)

# Convert to spatial data
cities_sf <- st_as_sf(colorado_cities, 
                      coords = c("longitude", "latitude"),
                      crs = 4326)
```

---

## Load Temperature Raster Data 

Load both temperature raster files using the `rast()` function that will be needed to complete this assignment. Adjust file paths if needed!

```{r}
# Load temperature raster data
co_temp_hist <- rast("data/co_historic_temp.tiff")
co_temp_2050 <- rast("data/co_2050_temp.tiff")

```

---

## Question 1: Visualize Historic Temperature (5 points)

Create a map showing historic temperature across Colorado. Your map should:
- Convert the raster to a data frame for plotting
- Use `geom_raster()` to display temperature values
- Add the Colorado state boundary as an overlay
- Include appropriate color scale and labels
- Use `scale_fill_distiller()` and the "RdYlBu" palette to make the colors more appropriate for viewing temperature data.

```{r}
# YOUR CODE HERE
```


---

## Question 2: Calculate Temperature Change (5 points)

Calculate the projected change in temperature between 2050 and historic. Call the new raster object "temp_change" and use `plot()` to view it.

```{r}
# YOUR CODE HERE
```

---

## Question 3: Extract Temperature Data for Cities (5 points)

Extract the projected temperature change values for all major Colorado cities using the `extract()` function and add this information as a new column in the `cities_sf` dataset:

```{r}
# YOUR CODE HERE
```


Create a map showing cities sized by their projected temperature change:

```{r}
# YOUR CODE HERE
```



Which cities are projected to experience the greatest temperature increases? List the top 3 and their approximate temperature changes.

*Your answer here:*

---

## Question 4: County Temperature Analysis (5 points)

Calculate the mean projected temperature change for each county using the `zonal()` function. Add the values as a new column to your `co_counties` data.

```{r}
# YOUR CODE HERE
```



Create a map showing counties colored by their projected temperature change:

```{r}
# YOUR CODE HERE
```



Which counties are projected to warm the most and least? Do you notice any geographic patterns (e.g., eastern plains vs. mountains vs. western slope)?

*Your answer here:*


---

## Bonus Task: Change Correlations (2 extra credit points)

Join the *counties_precip* dataset we created together in class with the similar *counties_temp* dataset (or whatever you named it) you created in this assignment and create a scatter plot that shows the relationship between change in precipitation and change in temperature. What trends do you see?

Notes:
- To use any of the dplyr join functions (such as `left_join()`) you will first need to convert *counties_precip* and your new *counties_temp* object from an sf object to a data frame. You can do this with the `st_drop_geometry()` function.

```{r bonus-task}
# Your bonus code here (if attempting) 

```


---
