---
title: "ASSIGNMENT_1_Intro"
output: word_document
---

To answer each question, put the code in a code chunk directly below each question. For this assignment you will need to load the `tidyverse` and `palmerpenguins` packages, and use the `penguins` dataset like we did in the lesson.

Load in necessary packages:
```{r}
library(tidyverse)
library(palmerpenguins)
```


Load in the `penguins` data with the line of code below:
```{r}
data("penguins")
```


#### Exercises


1.  Why don't the following lines of code work? Fix each one so the code runs.

```{r eval=FALSE}
myList["Fruit"]
```

```{r}
# CORRECT ANSWER HERE

```


```{r eval=FALSE}
penguins$flipper_lenght_mm
```

```{r}
# CORRECT ANSWER HERE

```


```{r eval=FALSE}
penguins[island=='Dream',]
```

```{r}
# CORRECT ANSWER HERE

```



2.  How many species are in the `penguins` data set? What islands were the data collected for? Note: the `unique()` function might help.


```{r}
# YOUR CODE HERE

```


3.  Use indexing to create a new data frame that has only 3 columns: species, island and flipper length columns, and subset all rows for just the 'Dream' island.

```{r}
# YOUR CODE HERE

```



4.  Use indexing and the `mean()` function to find the average flipper length for the *Adelie* species on Dream island. Note: explore the `mean()` function and how to deal with NA values.


```{r}
# YOUR CODE HERE

```



5.  Reorder the variables in `penguins` so that `year` is the first column followed by the rest. Hint: look into the use of `everything()`.


```{r}
# YOUR CODE HERE

```



6.  Create a new column called 'size_group' where individuals with a body mass greater than 4000 are called 'large' and those smaller are called 'small'.


```{r}
# YOUR CODE HERE

```




7.  Which species had the largest average body mass? Remember to account for NA values.


```{r}
# YOUR CODE HERE

```



8.  Filter data for years that are *not* 2008 or 2009.


```{r}
# YOUR CODE HERE

```



